---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "man/figures/README-",
                      dpi = 110)
options(width = 80)
library(drawn)
```

# drawn

<!-- badges: start -->
[![R-CMD-check](https://github.com/elkronos/dRawn/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/elkronos/dRawn/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**Take a probability sample, then say something defensible about the
population it came from.**

Pulling rows out of a data frame is easy. The hard part is what comes next:
someone asks what the total was across all 40,000 invoices, and you examined
200. Answering honestly needs to know how likely each examined row was to be
picked, and that number lives in the *design*, not in the sample. `drawn` keeps
the design around, so a sample can answer for itself:

```{r setup-frame}
set.seed(42)
invoices <- data.frame(
  id    = 1:600,
  site  = rep(c("north", "south", "east", "west"), times = c(300, 180, 90, 30)),
  value = round(rlnorm(600, meanlog = 6, sdlog = 1.1), 2)
)

plan <- design_stratified(strata = "site", n = 60, min_per_stratum = 2)
s <- draw(invoices, plan, seed = 1, weights = TRUE)

ht_total(s, "value")
```

The true total is `r format(sum(invoices$value), big.mark = ",")`. The
standard error is the point: without it you have a number; with it you have an
estimate, an interval on the design's own degrees of freedom, and a design
effect that says what the design bought you.

## Installation

```r
# install.packages("remotes")
remotes::install_github("elkronos/dRawn")
```

## What it does that other packages don't

R has excellent sampling software. `drawn` is built to sit between it, and to
make the step from "draw" to "estimate" hard to get wrong.

| | `drawn` | `sampling` | `survey` | `spsurvey`, `BalancedSampling` | `dplyr::slice_sample()` |
|---|---|---|---|---|---|
| Draws the sample | yes | yes | no | yes | yes |
| Data frame in, data frame out | yes | index vectors | — | yes / vectors | yes |
| Design is a reusable, printable value | yes | no | yes (for analysis) | no | no |
| Inclusion probabilities *before* drawing | yes, exact | yes | no | yes | no |
| Standard errors from the same object | yes | separate functions | yes | yes | no |
| Variance estimator chosen to match the design | yes, and named | you choose | you specify | yes (spatial) | — |
| Plan → draw → estimate → hand off | one package | — | — | — | — |

Where the others go further, use them: `sampling` has a dozen
unequal-probability algorithms and calibration; `survey` does regression,
calibration and replicate weights, and `as_svydesign()` hands a sample
straight to it; `spsurvey` has GRTS designs and a full environmental-monitoring
workflow.

## The workflow

**1. Size it.** You rarely know `n`; you know the margin of error you can live
with.

```{r plan}
plan_size(margin = 5, sd = 40, N = 20000, deff = 1.5, response = 0.8)
```

**2. Describe the design, and interrogate it before drawing.** A design is a
value. Ask it which rows it can reach, and at what probability:

```{r interrogate}
tapply(inclusion_prob(invoices, plan), invoices$site, unique)
```

**3. Draw,** with `weights = TRUE` to attach each row's inclusion probability
(`.prob`) and design weight (`.weight`).

**4. Check what you got.**

```{r summary}
sample_summary(s)
```

**5. Estimate** a total, a mean or a proportion, overall or by domain. Domain
estimates use the whole sample, not a subset of it, which is what keeps their
standard errors right:

```{r estimate}
ht_mean(s, "value", by = "site")
```

**6. Hand off** anything this package does not do:

```r
des <- as_svydesign(s)             # a survey.design, with matching SEs
survey::svyquantile(~value, des, 0.5)
```

## The designs

| Design | Selects | Size | Variance estimator |
|---|---|---|---|
| `design_simple()` | rows uniformly at random | `n` | exact |
| `design_stratified()` | a share of each stratum: proportional, equal or Neyman | `n` total | exact |
| `design_systematic()` | every *k*-th row from a random start | `interval` | successive-difference approx. |
| `design_cluster()` | whole clusters | `n_clusters` | exact |
| `design_multistage()` | clusters, then rows within them | `n_clusters`, `n` | exact two-stage |
| `design_weighted()` | probability proportional to size | `n` | exact (Poisson), Deville (systematic) |
| `design_certainty()` | everything above a threshold, plus any design below it | `rest`'s own | `rest`'s own |
| `design_spread()` | a sample spread evenly over a map or covariates | `n` | local-mean approx. |
| `design_temporal()` | a fixed number per time interval | `per_interval` | exact |
| `design_spatial()` | a simple random sample inside a region | `n` | exact |
| `design_reservoir()` | a fixed-size sample from a stream, in one pass | `n` | exact |
| `design_bootstrap()` | resampled replicates, simple or block | `n_replicates` | not a survey design |

Every estimate says which estimator produced its variance, and any
approximation is named when it prints.

## Spread the sample, not just the rows

When what you measure varies smoothly across space — biomass, soil carbon,
footfall — a sample that covers the map evenly is worth far more than one that
clumps. `design_spread()` uses the local pivotal method (Grafström, Lundström
and Schelin 2012): exact inclusion probabilities, but neighbours compete, so
clumps and gaps become rare.

```{r spread, fig.width = 8, fig.height = 3.6, fig.show = "hold", fig.alt = "Two maps of 400 plots. On the left, a simple random sample of 40 leaves clumps and empty patches; on the right, the spread sample covers the square evenly."}
set.seed(1)
plots <- data.frame(x = runif(400), y = runif(400))
plots$biomass <- 100 + 80 * plots$x + 60 * sin(4 * plots$y) + rnorm(400, 0, 10)

op <- par(mfrow = c(1, 2))
plot(design_simple(n = 40), plots, type = "map", coords = c("x", "y"), seed = 3,
     main = "simple random")
plot(design_spread(c("x", "y"), n = 40, scale = FALSE), plots, type = "map",
     seed = 3, main = "spread (local pivotal method)")
par(op)
```

```{r spread-gain}
spread_of <- function(design) sd(vapply(1:200, function(i) {
  ht_total(draw(plots, design, seed = i, weights = TRUE), "biomass",
           variance = "none")$total
}, numeric(1)))

c(simple = spread_of(design_simple(n = 40)),
  spread = spread_of(design_spread(c("x", "y"), n = 40, scale = FALSE)))
```

Same sample size, a fraction of the error.

## Told when the answer isn't available

Some designs have no closed-form inclusion probability, and some samples cannot
support a variance. `drawn` says so, and says what to change, rather than
returning a plausible number:

```{r refuse, error = TRUE}
inclusion_prob(invoices, design_weighted("value", n = 60))
```

```{r lonely}
one_each <- draw(invoices, design_stratified("site", n = 4, min_per_stratum = 1),
                 seed = 1, weights = TRUE)
ht_total(one_each, "value")
```

## Learn more

* [Get started](https://elkronos.github.io/dRawn/articles/drawn.html) walks
  through the whole workflow on one frame.
* [Choosing a design](https://elkronos.github.io/dRawn/articles/choosing-a-design.html)
  compares the designs on the same population, and shows when each one pays.
* [How the estimates work](https://elkronos.github.io/dRawn/articles/estimation.html)
  sets out every variance estimator, with references and simulation checks.
* Walkthroughs: [an audit sample](https://elkronos.github.io/dRawn/articles/audit-sampling.html)
  and [a spatially balanced field survey](https://elkronos.github.io/dRawn/articles/spatial-sampling.html).

## Credit

The sampling routines began as an R port of
[sample_py](https://github.com/elkronos/sample_py).

## License

GPL-3
