---
title: "Choosing a sampling design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a sampling design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(drawn)
```

Every function in this package answers one question — *which rows do I look at?*
— and the interesting part is what you can say afterwards about the rows you
didn't.

## One population, eleven questions

We'll use a synthetic audit population throughout: invoices from a handful of
sites, over a fortnight, with a value attached.

```{r}
set.seed(42)
n <- 600
invoices <- data.frame(
  id    = seq_len(n),
  site  = rep(c("north", "south", "east", "west"), times = c(300, 180, 90, 30)),
  team  = rep(paste0("t", 1:20), each = 30),
  when  = rep(seq(as.POSIXct("2024-03-01", tz = "UTC"), by = "day",
                  length.out = 15), each = 40),
  value = round(stats::rlnorm(n, meanlog = 6, sdlog = 1.1), 2)
)

# Two things we could only learn by opening the invoice: what the review cost
# (scales with size) and how many checks it needed (doesn't)
invoices$cost <- round(invoices$value * stats::runif(n, 0.02, 0.06), 2)
invoices$checks <- stats::rpois(n, 3) + 1

str(invoices)
sum(invoices$value)
```

The population total is what we'll try to recover from a sample of 60.

## How many rows? Ask before you choose a design

Sixty is a stand-in for a number you should really derive. `plan_size()` inverts
the usual question: rather than guessing `n` and hoping, say what margin of
error you can live with and let it solve for the size.

```{r}
plan_size(margin = 100, sd = stats::sd(invoices$value), N = nrow(invoices))
```

The frame size matters — `N` applies a finite population correction, and 600
invoices is small enough that it bites hard. So do two other facts about the
real world:

```{r}
# A clustered design carries less information per row, and people don't respond
plan_size(margin = 100, sd = stats::sd(invoices$value), N = nrow(invoices),
          deff = 2.5, response = 0.7)
```

`deff` is not a guess you have to invent — it is measurable from a comparable
past sample with `deff()`, which we come back to below. If the margin you ask
for cannot be reached by sampling at all, `plan_size()` says so rather than
returning a number bigger than your frame:

```{r}
plan_size(margin = 2, sd = stats::sd(invoices$value), N = nrow(invoices))
```

We'll stay with 60 for the rest of this vignette, because it makes the
comparisons between designs legible.

## A design is a value, not a function call

Designs are built, then applied. That separation is the point: the same design
can be printed, stored, reused on next month's data, and — the part that
matters — asked what it does.

```{r}
plan <- design_stratified(strata = "site", n = 60)
plan

sample_1 <- draw(invoices, plan, seed = 1)
table(sample_1$site)
```

Proportional allocation splits 60 across sites in proportion to their size, and
the total is exact — no rounding drift.

```{r}
nrow(sample_1)
```

## Estimating from the sample

A sample is only useful for estimation if you know how likely each row was to be
in it. `weights = TRUE` attaches that.

```{r}
s <- draw(invoices, plan, seed = 1, weights = TRUE)
head(s[, c(".prob", ".weight", "site", "value")], 4)
```

`.weight` is `1 / .prob`: the number of population rows each sampled row stands
for. Multiply and add up, and you have a Horvitz–Thompson total.

```{r}
ht <- sum(s$value * s$.weight)
c(estimate = round(ht), truth = sum(invoices$value))
```

One draw won't land on the truth. `ht_total()` gives you the total *and* a
standard error, so you know how far off it might be:

```{r}
ht_total(s, "value")
```

That interval comes from second-order inclusion probabilities — the chance a
*pair* of rows both land in the sample — which `joint_prob()` computes:

```{r}
round(joint_prob(invoices, plan, rows = 1:4), 4)
```

The estimator is unbiased — the average over many
draws converges.

```{r}
ests <- vapply(1:400, function(i) {
  si <- draw(invoices, plan, seed = i, weights = TRUE)
  sum(si$value * si$.weight)
}, numeric(1))

c(mean_estimate = round(mean(ests)), truth = sum(invoices$value))
```

And the reported standard error tracks the estimator's real spread:

```{r}
ses <- vapply(1:400, function(i) {
  ht_total(draw(invoices, plan, seed = i, weights = TRUE), "value")$se
}, numeric(1))

c(mean_reported_se = round(mean(ses)), actual_sd = round(stats::sd(ests)))
```

You can also ask the design about the population *before* drawing anything,
which is a quick way to check a plan is doing what you meant:

```{r}
p <- inclusion_prob(invoices, plan)
tapply(p, invoices$site, unique)
```

Every site is sampled at very nearly 10%, which is what proportional allocation
means.

## Means, and what the design cost you

`ht_mean()` is the same machinery pointed at an average instead of a total:

```{r}
ht_mean(s, "value")
mean(invoices$value)
```

It defaults to the Hájek estimator, `sum(y/pi) / sum(1/pi)`, which divides by
the *estimated* population size rather than the known one. `estimator = "ht"`
divides by the true `N`. The two agree exactly whenever the weights of the rows
you drew sum to `N` — which includes the stratified design above, so here they
are the same number:

```{r}
c(hajek = ht_mean(s, "value", variance = "none")$mean,
  ht    = ht_mean(s, "value", "ht", variance = "none")$mean)
```

They part company when the sample size is random. Poisson sampling is the clean
case — and which of the two wins depends on what you are averaging:

```{r}
pois <- design_weighted("value", n = 60, method = "poisson")

spread_both <- function(column) {
  b <- vapply(1:300, function(i) {
    si <- draw(invoices, pois, seed = i, weights = TRUE)
    c(hajek = ht_mean(si, column, variance = "none")$mean,
      ht    = ht_mean(si, column, "ht", variance = "none")$mean)
  }, numeric(2))
  apply(b, 1, stats::sd)
}

round(sapply(c(checks = "checks", cost = "cost", value = "value"),
             spread_both), 2)
```

Selection here is driven by `value`. For `checks`, which has nothing to do with
invoice size, Hájek is more than twice as steady — a draw that happens to catch
too many low-probability rows inflates numerator and denominator together, and
they partly cancel. For `cost`, which scales with size, and for `value` itself,
that cancellation is exactly what you *don't* want: `y / pi` is already nearly
constant, so dividing by the known `N` wins.

Hájek is the default because the first case is the common one — you usually
select on size and measure something else. But it is a tendency, not a law, and
this is how you check which side of it you are on.

Every estimate also reports a **design effect**: the variance under this design
against simple random sampling of the same size. Read it as an exchange rate on
sample size — at `deff = 2`, a sample of 400 carries the information of 200.

To see it move, we need a frame where the grouping variable genuinely matters.
Regions on very different value levels, and delivery routes that each sit inside
one region:

```{r}
set.seed(11)
regions <- data.frame(
  id     = 1:400,
  region = rep(c("north", "south", "east", "west"), each = 100),
  route  = rep(paste0("r", 1:40), each = 10)
)
regions$value <- rep(c(120, 260, 480, 900), each = 100) +
  round(stats::rnorm(400, 0, 40))

by_region <- draw(regions, design_stratified("region", n = 40), seed = 1,
                  weights = TRUE)
by_route <- draw(regions, design_cluster("route", n_clusters = 4), seed = 1,
                 weights = TRUE)

c(stratified = deff(ht_mean(by_region, "value")),
  clustered  = deff(ht_mean(by_route,  "value")))
```

Same population, same forty rows. Stratifying on what varies removes nearly all
the variance; clustering on it keeps nearly all of it. That is the range a
design choice moves you across — and that number is exactly what
`plan_size(deff = )` wants for the next study, which closes the loop.

## Checking what you actually got

Before analysing, look at what came back against what was there:

```{r}
sample_summary(s)
```

This one is healthy: every stratum came back at its intended 10%, and the
weights are constant. That is what you want to confirm before analysing, because
the three things that quietly ruin an estimate all show up here — strata that
came up short, design weights that vary far more than expected, and rows the
design could never have reached at all. Temporal designs are the usual source of
that last one, since anything outside the sampling window has probability zero:

```{r}
window <- design_temporal("when", from = "2024-03-01", to = "2024-03-08",
                          interval = 1, per_interval = 4, unit = "days")
sample_summary(draw(invoices, window, seed = 1, weights = TRUE))
```

## When you want the rare stratum represented

`west` has only 30 invoices. Proportional allocation gives it three. If you need
coverage rather than efficiency, ask for a floor — and note that this is a
deliberate bias, not a free lunch:

```{r}
covered <- draw(invoices, design_stratified("site", n = 60, min_per_stratum = 8),
                seed = 1)
table(covered$site)
```

The default of `min_per_stratum = 0` leaves allocation unbiased. Setting it
over-represents small strata, which is fine when the goal is "look at every
site" and wrong when the goal is "estimate the total".

## Sampling proportional to size

For an audit you usually want large invoices sampled more often. The obvious
call is not quite the right one:

```{r}
naive <- design_weighted("value", n = 60)
naive
```

This is what `base::sample(prob = )` does. It's a perfectly good way to bias
selection toward large invoices, but the weights govern each sequential *draw*,
not the probability of ending up in the sample — so it has no closed-form
inclusion probability, and the package refuses to invent one:

```{r, error = TRUE}
inclusion_prob(invoices, naive)
```

Ask for a design whose inclusion probabilities really are proportional to size:

```{r}
pps <- design_weighted("value", n = 60, method = "systematic")
pi <- inclusion_prob(invoices, pps)
round(head(pi / invoices$value, 5), 8)
```

A constant ratio to `value` is what "proportional to size" means. And now the
estimator works:

```{r}
ests_pps <- vapply(1:400, function(i) {
  si <- draw(invoices, pps, seed = i, weights = TRUE)
  sum(si$value * si$.weight)
}, numeric(1))

c(mean_estimate = round(mean(ests_pps)), truth = sum(invoices$value))
```

Compare its spread against the stratified estimator's, and something surprising
happens:

```{r}
c(stratified_sd = round(stats::sd(ests)),
  pps_sd        = round(stats::sd(ests_pps)))
```

Zero. Not "very small" — exactly zero, in every sample. That is not a fluke, it
is the limiting case, and it is worth understanding because it says exactly what
PPS is for. Under systematic PPS the inclusion probability is
`pi_i = n * value_i / sum(value)`, so each sampled row contributes
`value_i / pi_i = sum(value) / n` — the same amount, whichever rows you happen
to draw. Totalling the very variable that drove selection gives the answer back
exactly.

Real work is never that kind. You select on a size measure you already hold for
the whole frame, and you total something you can only observe once you open the
file. Whether PPS helps then depends entirely on whether the two are related:

```{r}
spread_of <- function(design, column) {
  stats::sd(vapply(1:400, function(i) {
    si <- draw(invoices, design, seed = i, weights = TRUE)
    sum(si[[column]] * si$.weight)
  }, numeric(1)))
}

round(rbind(
  cost   = c(stratified = spread_of(plan, "cost"),
             pps        = spread_of(pps,  "cost")),
  checks = c(stratified = spread_of(plan, "checks"),
             pps        = spread_of(pps,  "checks"))
), 1)
```

Review cost scales with invoice size, and PPS estimates its total several times
more precisely than stratification does. The number of checks an invoice needs
does not scale with size — and there PPS is markedly *worse*, because it spends
most of the sample on large invoices that carry no more information about checks
than small ones do.

That is the trade PPS actually offers. It is a bet that your size measure
predicts what you are measuring, and it is a bad bet when it doesn't. `deff()`,
below, puts a number on which way the bet went.

## The units too big to leave to chance

PPS is one answer to dominant units. The blunter one is to stop sampling them
at all: take every invoice above a threshold, and sample what is left.

```{r}
audit <- design_certainty("value", threshold = 2000,
                          rest = design_stratified("site", n = 60))
audit

s_audit <- draw(invoices, audit, seed = 1, weights = TRUE)
table(certain = s_audit$.prob == 1)
```

`rest` takes any design, and its `n` counts rows drawn from *below* the
threshold, not the total. The rows above it have inclusion probability exactly
1 — so they carry a weight of 1, contribute their own value to the total, and
add **nothing** to its variance:

```{r}
ht_total(s_audit, "value")
```

All the uncertainty comes from the part that was actually sampled. That is the
whole point, and it is why this is the standard shape of an audit or financial
sample. Compare it against stratifying alone on the same frame:

```{r}
c(stratified = ht_total(s, "value")$se,
  certainty  = ht_total(s_audit, "value")$se)
```

Inclusion and joint probabilities compose exactly across the two parts — a
certainty row co-occurs with another row exactly when that row is drawn, and two
certainty rows always co-occur — so that standard error is analytic, not an
approximation.

## The designs at a glance

```{r}
designs <- list(
  simple      = design_simple(n = 60),
  stratified  = design_stratified("site", n = 60),
  systematic  = design_systematic(interval = 10),
  cluster     = design_cluster("team", n_clusters = 4),
  multistage  = design_multistage("team", n_clusters = 6, n = 60),
  weighted    = design_weighted("value", n = 60, method = "systematic"),
  certainty   = design_certainty("value", 2000, design_simple(n = 60)),
  temporal    = design_temporal("when", from = "2024-03-01", to = "2024-03-15",
                                interval = 1, per_interval = 4, unit = "days")
)

vapply(designs, function(d) nrow(draw(invoices, d, seed = 1)), numeric(1))
```

`design_cluster()` has no `n`: you choose how many clusters, and the row count
follows from which ones you get. Use `design_multistage()` when you need to
control both.

## Which designs can be estimated from

Most designs have exact inclusion probabilities. Five situations do not, and the
package says so rather than approximating:

```{r}
no_form <- list(
  `balanced clusters`      = design_cluster("team", n_clusters = 4, balanced = TRUE),
  `proportional multistage`= design_multistage("team", n_clusters = 6, n = 60,
                                               allocation = "proportional"),
  `uneven multistage`      = design_multistage("team", n_clusters = 6, n = 61),
  `successive weighting`   = design_weighted("value", n = 60),
  `bootstrap`              = design_bootstrap(n_replicates = 10)
)

for (nm in names(no_form)) {
  msg <- tryCatch(inclusion_prob(invoices, no_form[[nm]]),
                  error = function(e) conditionMessage(e))
  cat(nm, "->", strsplit(msg, "\n")[[1]][1], "\n")
}
```

In each case the probability depends on something random — which clusters were
picked, or the order draws happened in. The multistage pair is worth dwelling
on, because the arithmetic looks innocent. The second-stage allocation runs over
the clusters that were *selected*, so a leftover row goes to the largest of
them, and a cluster smaller than its target is capped with the shortfall dealt
to whichever clusters came with it. A row's chance therefore depends on the
company it keeps. Averaging over the possibilities is not a closed form, and the
average is not the answer.

Where you need a number anyway, ask for a simulation and accept the Monte Carlo
error:

```{r}
sim <- inclusion_prob(invoices, no_form$`balanced clusters`,
                      simulate = TRUE, R = 200, seed = 1)
round(tapply(sim, invoices$team, unique)[1:5], 3)
```

The one thing simulation cannot rescue is the bootstrap, and it is refused
rather than answered — every row turns up in some replicate, so the count would
converge to 1 for all of them.

```{r, error = TRUE}
inclusion_prob(invoices, no_form$bootstrap, simulate = TRUE, R = 100)
```

Second-order probabilities can be simulated too, which is the general answer
where no formula exists. Systematic PPS is the case that needs it: it has exact
first-order probabilities, but its joint ones depend on the order units are
visited.

```{r, error = TRUE}
joint_prob(invoices, pps, rows = 1:4)
```

```{r}
round(joint_prob(invoices, pps, rows = 1:4, simulate = TRUE, R = 2000,
                 seed = 1), 4)
```

The diagonal holds the first-order probabilities, and simulating them is a
useful check on the closed form:

```{r}
round(cbind(
  simulated = diag(joint_prob(invoices, pps, rows = 1:4, simulate = TRUE,
                              R = 2000, seed = 1)),
  exact     = inclusion_prob(invoices, pps)[1:4]
), 4)
```

## Clusters, when fieldwork costs more than precision

Visiting five teams is cheaper than visiting twenty. `design_cluster()` takes
whole clusters, so the row count follows from which ones you get:

```{r}
by_team <- design_cluster("team", n_clusters = 4)
res <- draw(invoices, by_team, seed = 1)
c(teams = length(unique(res$team)), rows = nrow(res))
```

`design_multistage()` adds a second stage when you need to control the total:

```{r}
ms <- design_multistage("team", n_clusters = 6, n = 60)
res_ms <- draw(invoices, ms, seed = 1)
table(res_ms$team)
```

Both are estimable, but cluster designs pay for their convenience in precision —
and in degrees of freedom. With only a handful of clusters the normal-approximation
interval undercovers, so treat it as indicative:

```{r}
ht_total(draw(invoices, by_team, seed = 1, weights = TRUE), "value")
```

## Streams, when the data does not fit

A data frame is not a stream: its length is already known, so `design_reservoir()`
takes a direct vectorised path for one. The streaming path is for data you cannot
hold — pass a connection, or a function that returns the next item and `NULL` when
exhausted.

```{r}
i <- 0
next_invoice <- function() {
  i <<- i + 1
  if (i > 1e5) NULL else i
}

drawn_ids <- unlist(draw(next_invoice, design_reservoir(n = 5), seed = 1))
drawn_ids
```

One pass, five items held, a hundred thousand seen. Algorithm L skips ahead
geometrically rather than drawing a random number per item.

## Bootstrap, for the uncertainty of a statistic

The other designs estimate a population total. The bootstrap estimates the
sampling distribution of whatever you like — here the median, which has no
convenient closed form:

```{r}
reps <- draw(invoices, design_bootstrap(n_replicates = 400), seed = 1)
meds <- vapply(split(reps, reps$.replicate), function(r) median(r$value),
               numeric(1))

c(observed = median(invoices$value),
  boot_se  = round(stats::sd(meds), 1),
  lower    = round(unname(stats::quantile(meds, 0.025)), 1),
  upper    = round(unname(stats::quantile(meds, 0.975)), 1))
```

Replicates come back in one frame with a leading `.replicate` column. For ordered
data where neighbouring rows are correlated, `method = "block"` concatenates
randomly chosen runs instead of independent rows:

```{r}
design_bootstrap(n_replicates = 100, method = "block", block_length = 20)
```

## Reproducibility without side effects

`seed` is local to the draw. Sampling inside a simulation does not shift the
simulation's own random stream:

```{r}
set.seed(99)
before <- runif(3)

set.seed(99)
invisible(draw(invoices, plan, seed = 7))
after <- runif(3)

identical(before, after)
```

That is deliberate: a sampling call is not supposed to be visible to the code
around it.

## Where this package stops

`drawn` draws samples and estimates totals and means from them. It does not do
subpopulation estimates, regression, calibration, or quantiles with proper
standard errors — [`survey`](https://cran.r-project.org/package=survey) does all
of that, and `as_svydesign()` hands a sample straight over.

```{r, eval = requireNamespace("survey", quietly = TRUE), message = FALSE}
library(survey)

des <- as_svydesign(s)
svytotal(~value, des)
```

The mapping is direct: design weights become `weights`, the stratification or
clustering column becomes `strata` or `ids`, and the population counts become
`fpc` — so `survey` applies the same finite population correction this package
does, and the two agree.

```{r, eval = requireNamespace("survey", quietly = TRUE)}
c(drawn  = ht_total(s, "value")$se,
  survey = as.numeric(SE(svytotal(~value, des))))
```

From there it is `survey`'s vocabulary:

```{r, eval = requireNamespace("survey", quietly = TRUE)}
svyby(~value, ~site, des, svymean)
```

Draw here, analyse there.
