---
title: "Choosing a sampling design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a sampling design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(drawn)
```

Every function in this package answers one question — *which rows do I look at?*
— and the interesting part is what you can say afterwards about the rows you
didn't.

## One population, ten questions

We'll use a synthetic audit population throughout: invoices from a handful of
sites, over a fortnight, with a value attached.

```{r}
set.seed(42)
n <- 600
invoices <- data.frame(
  id    = seq_len(n),
  site  = rep(c("north", "south", "east", "west"), times = c(300, 180, 90, 30)),
  team  = rep(paste0("t", 1:20), each = 30),
  when  = rep(seq(as.POSIXct("2024-03-01", tz = "UTC"), by = "day",
                  length.out = 15), each = 40),
  value = round(stats::rlnorm(n, meanlog = 6, sdlog = 1.1), 2)
)

str(invoices)
sum(invoices$value)
```

The population total is what we'll try to recover from a sample of 60.

## A design is a value, not a function call

Designs are built, then applied. That separation is the point: the same design
can be printed, stored, reused on next month's data, and — the part that
matters — asked what it does.

```{r}
plan <- design_stratified(strata = "site", n = 60)
plan

sample_1 <- draw(invoices, plan, seed = 1)
table(sample_1$site)
```

Proportional allocation splits 60 across sites in proportion to their size, and
the total is exact — no rounding drift.

```{r}
nrow(sample_1)
```

## Estimating from the sample

A sample is only useful for estimation if you know how likely each row was to be
in it. `weights = TRUE` attaches that.

```{r}
s <- draw(invoices, plan, seed = 1, weights = TRUE)
head(s[, c(".prob", ".weight", "site", "value")], 4)
```

`.weight` is `1 / .prob`: the number of population rows each sampled row stands
for. Multiply and add up, and you have a Horvitz–Thompson total.

```{r}
ht <- sum(s$value * s$.weight)
c(estimate = round(ht), truth = sum(invoices$value))
```

One draw won't land on the truth. `ht_total()` gives you the total *and* a
standard error, so you know how far off it might be:

```{r}
ht_total(s, "value")
```

That interval comes from second-order inclusion probabilities — the chance a
*pair* of rows both land in the sample — which `joint_prob()` computes:

```{r}
round(joint_prob(invoices, plan, rows = 1:4), 4)
```

The estimator is unbiased — the average over many
draws converges.

```{r}
ests <- vapply(1:400, function(i) {
  si <- draw(invoices, plan, seed = i, weights = TRUE)
  sum(si$value * si$.weight)
}, numeric(1))

c(mean_estimate = round(mean(ests)), truth = sum(invoices$value))
```

And the reported standard error tracks the estimator's real spread:

```{r}
ses <- vapply(1:400, function(i) {
  ht_total(draw(invoices, plan, seed = i, weights = TRUE), "value")$se
}, numeric(1))

c(mean_reported_se = round(mean(ses)), actual_sd = round(stats::sd(ests)))
```

You can also ask the design about the population *before* drawing anything,
which is a quick way to check a plan is doing what you meant:

```{r}
p <- inclusion_prob(invoices, plan)
tapply(p, invoices$site, unique)
```

Every site is sampled at very nearly 10%, which is what proportional allocation
means.

## When you want the rare stratum represented

`west` has only 30 invoices. Proportional allocation gives it three. If you need
coverage rather than efficiency, ask for a floor — and note that this is a
deliberate bias, not a free lunch:

```{r}
covered <- draw(invoices, design_stratified("site", n = 60, min_per_stratum = 8),
                seed = 1)
table(covered$site)
```

The default of `min_per_stratum = 0` leaves allocation unbiased. Setting it
over-represents small strata, which is fine when the goal is "look at every
site" and wrong when the goal is "estimate the total".

## Sampling proportional to size

For an audit you usually want large invoices sampled more often. The obvious
call is not quite the right one:

```{r}
naive <- design_weighted("value", n = 60)
naive
```

This is what `base::sample(prob = )` does. It's a perfectly good way to bias
selection toward large invoices, but the weights govern each sequential *draw*,
not the probability of ending up in the sample — so it has no closed-form
inclusion probability, and the package refuses to invent one:

```{r, error = TRUE}
inclusion_prob(invoices, naive)
```

Ask for a design whose inclusion probabilities really are proportional to size:

```{r}
pps <- design_weighted("value", n = 60, method = "systematic")
pi <- inclusion_prob(invoices, pps)
round(head(pi / invoices$value, 5), 8)
```

A constant ratio to `value` is what "proportional to size" means. And now the
estimator works:

```{r}
ests_pps <- vapply(1:400, function(i) {
  si <- draw(invoices, pps, seed = i, weights = TRUE)
  sum(si$value * si$.weight)
}, numeric(1))

c(mean_estimate = round(mean(ests_pps)), truth = sum(invoices$value))
```

Because selection probability tracks the thing being totalled, the PPS estimator
is also far more precise than the stratified one here:

```{r}
c(stratified_sd = round(stats::sd(ests)),
  pps_sd        = round(stats::sd(ests_pps)))
```

## The designs at a glance

```{r}
designs <- list(
  simple      = design_simple(n = 60),
  stratified  = design_stratified("site", n = 60),
  systematic  = design_systematic(interval = 10),
  cluster     = design_cluster("team", n_clusters = 4),
  multistage  = design_multistage("team", n_clusters = 6, n = 60),
  weighted    = design_weighted("value", n = 60, method = "systematic"),
  temporal    = design_temporal("when", from = "2024-03-01", to = "2024-03-15",
                                interval = 1, per_interval = 4, unit = "days")
)

vapply(designs, function(d) nrow(draw(invoices, d, seed = 1)), numeric(1))
```

`design_cluster()` has no `n`: you choose how many clusters, and the row count
follows from which ones you get. Use `design_multistage()` when you need to
control both.

## Which designs can be estimated from

Nine of the ten have exact inclusion probabilities. Four situations do not, and
the package says so rather than approximating:

```{r}
no_form <- list(
  `balanced clusters`      = design_cluster("team", n_clusters = 4, balanced = TRUE),
  `proportional multistage`= design_multistage("team", n_clusters = 6, n = 60,
                                               allocation = "proportional"),
  `successive weighting`   = design_weighted("value", n = 60),
  `bootstrap`              = design_bootstrap(n_replicates = 10)
)

for (nm in names(no_form)) {
  msg <- tryCatch(inclusion_prob(invoices, no_form[[nm]]),
                  error = function(e) conditionMessage(e))
  cat(nm, "->", strsplit(msg, "\n")[[1]][1], "\n")
}
```

In each case the reason is that the probability depends on something random —
which clusters were picked, or the order draws happened in. Where you need a
number anyway, ask for a simulation and accept the Monte Carlo error:

```{r}
sim <- inclusion_prob(invoices, no_form$`balanced clusters`,
                      simulate = TRUE, R = 200, seed = 1)
round(tapply(sim, invoices$team, unique)[1:5], 3)
```

## Clusters, when fieldwork costs more than precision

Visiting five teams is cheaper than visiting twenty. `design_cluster()` takes
whole clusters, so the row count follows from which ones you get:

```{r}
by_team <- design_cluster("team", n_clusters = 4)
res <- draw(invoices, by_team, seed = 1)
c(teams = length(unique(res$team)), rows = nrow(res))
```

`design_multistage()` adds a second stage when you need to control the total:

```{r}
ms <- design_multistage("team", n_clusters = 6, n = 60)
res_ms <- draw(invoices, ms, seed = 1)
table(res_ms$team)
```

Both are estimable, but cluster designs pay for their convenience in precision —
and in degrees of freedom. With only a handful of clusters the normal-approximation
interval undercovers, so treat it as indicative:

```{r}
ht_total(draw(invoices, by_team, seed = 1, weights = TRUE), "value")
```

## Streams, when the data does not fit

A data frame is not a stream: its length is already known, so `design_reservoir()`
takes a direct vectorised path for one. The streaming path is for data you cannot
hold — pass a connection, or a function that returns the next item and `NULL` when
exhausted.

```{r}
i <- 0
next_invoice <- function() {
  i <<- i + 1
  if (i > 1e5) NULL else i
}

drawn_ids <- unlist(draw(next_invoice, design_reservoir(n = 5), seed = 1))
drawn_ids
```

One pass, five items held, a hundred thousand seen. Algorithm L skips ahead
geometrically rather than drawing a random number per item.

## Bootstrap, for the uncertainty of a statistic

The other designs estimate a population total. The bootstrap estimates the
sampling distribution of whatever you like — here the median, which has no
convenient closed form:

```{r}
reps <- draw(invoices, design_bootstrap(n_replicates = 400), seed = 1)
meds <- vapply(split(reps, reps$.replicate), function(r) median(r$value),
               numeric(1))

c(observed = median(invoices$value),
  boot_se  = round(stats::sd(meds), 1),
  lower    = round(unname(stats::quantile(meds, 0.025)), 1),
  upper    = round(unname(stats::quantile(meds, 0.975)), 1))
```

Replicates come back in one frame with a leading `.replicate` column. For ordered
data where neighbouring rows are correlated, `method = "block"` concatenates
randomly chosen runs instead of independent rows:

```{r}
design_bootstrap(n_replicates = 100, method = "block", block_length = 20)
```

## Reproducibility without side effects

`seed` is local to the draw. Sampling inside a simulation does not shift the
simulation's own random stream:

```{r}
set.seed(99)
before <- runif(3)

set.seed(99)
invisible(draw(invoices, plan, seed = 7))
after <- runif(3)

identical(before, after)
```

That is deliberate: a sampling call is not supposed to be visible to the code
around it.
