---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.show = "hold", fig.width = 7,
                      fig.height = 3.4, dpi = 96)
```

```{r setup}
library(drawn)
```

This walkthrough follows one job from start to finish: a retail chain wants to
know how much stock it lost last quarter across its 2,000 stores, and can afford
to audit a sample of them. Along the way it covers the six steps every use of
`drawn` goes through — size, design, check, draw, estimate, hand off.

## The frame

What we know about every store *before* sampling: its region, its floor area,
and what it lost the year before. What we want — this quarter's loss — is only
known for the stores we audit. Here the frame holds it too, so we can check the
answers, but nothing below uses it except through the sample.

```{r frame}
set.seed(2024)
N <- 2000
stores <- data.frame(
  store  = sprintf("S%04d", 1:N),
  region = sample(c("north", "south", "east", "west", "central"), N,
                  replace = TRUE, prob = c(0.3, 0.25, 0.2, 0.15, 0.1)),
  area   = round(rlnorm(N, 7, 0.5))
)
region_rate <- c(north = 1.0, south = 1.4, east = 0.8, west = 1.9,
                 central = 1.1)
stores$last_year <- round(stores$area * region_rate[stores$region] *
                            rgamma(N, 4, 4) * 2)
stores$loss <- round(stores$last_year * rgamma(N, 12, 12) +
                       rnorm(N, 0, 200))
stores$loss <- pmax(stores$loss, 0)

# A size band from last year's loss: known for every store before sampling
stores$band <- cut(stores$last_year,
                   quantile(stores$last_year, c(0, 0.5, 0.8, 0.95, 1)),
                   labels = c("small", "medium", "large", "top"),
                   include.lowest = TRUE)

head(stores)
```

## 1. How many stores?

Last year's losses give a fair idea of the spread, and management wants the
total to within about 10%. `plan_size()` turns that into a number of stores —
allowing for audits that fail to complete, and assuming no help from the design
yet (`deff = 1`), which errs on the safe side:

```{r plan}
target <- 0.10 * sum(stores$last_year)
p <- plan_size(margin = target, sd = sd(stores$last_year), N = N,
               target = "total", response = 0.95)
p
```

## 2. Choose a design, and ask it questions before drawing

Losses track store size far more than region, and last year's loss is known
for every store, so stratify on the size band built from it. Allocate by
Neyman — more audits where losses vary most, which is among the biggest
stores — and set `min_per_stratum = 2` so every band is reachable and
contributes to the variance.

```{r design}
plan <- design_stratified("band", n = p$n, allocation = "neyman",
                          allocation_by = "last_year", min_per_stratum = 2)
plan
```

A design is a value. Before drawing anything, ask it what each store's chance
of selection is:

```{r interrogate}
pi <- inclusion_prob(stores, plan)
round(tapply(pi, stores$band, unique), 3)
```

The top band, where losses are largest and most variable, is sampled far more
heavily than the small stores. That is Neyman allocation doing its job, and the
design weights scale those stores back down in every estimate.

```{r plot-prob, fig.alt = "Inclusion probability for each of the 2,000 stores, sorted by size band: a rising step pattern with four levels."}
plot(plan, stores[order(stores$band), ], type = "probability")
```

## 3. Draw

`weights = TRUE` attaches each audited store's inclusion probability (`.prob`)
and design weight (`.weight`, the number of stores it stands for), and
remembers the design so the sample can be estimated from later. `seed` makes the
draw reproducible without disturbing your own random number stream.

```{r draw}
audit <- draw(stores, plan, seed = 7, weights = TRUE)
nrow(audit)
head(audit[, c(".prob", ".weight", "store", "band", "region", "loss")])
```

## 4. Check what you got

```{r check}
sample_summary(audit)
```

Every band is represented, no store is unreachable, and the weights vary by
band exactly as the allocation intended.

## 5. Estimate

The total loss, with a standard error and a confidence interval:

```{r total}
ht_total(audit, "loss")
sum(stores$loss)   # what we are trying to recover
```

The interval uses Student's t on the design's degrees of freedom — audited
stores minus strata — and the design effect says what the design bought: at
`r signif(deff(ht_total(audit, "loss")), 2)`, this sample carries about as much
information as `r round(nrow(audit) / deff(ht_total(audit, "loss")))` stores
drawn at random.

The mean loss per store, and the share of stores losing more than 1,500, are the
same machinery; a proportion is just the mean of a yes/no column:

```{r mean-prop}
ht_mean(audit, "loss")
ht_mean(audit, audit$loss > 1500)
```

Estimates for each region come from the whole sample, not from cutting it into
pieces — region was not a stratum, so how many audited stores land in each is
itself random, and the standard errors account for that:

```{r by}
ht_total(audit, "loss", by = "region")
```

## What the design bought

Drawing the same number of stores many times under each design shows what the
stratification was worth — and that stratifying on region, the obvious choice,
would have bought almost nothing:

```{r compare}
spread_of <- function(design) {
  sd(vapply(1:200, function(i) {
    ht_total(draw(stores, design, seed = i, weights = TRUE), "loss",
             variance = "none")$total
  }, numeric(1)))
}
round(c(simple    = spread_of(design_simple(n = p$n)),
        by_region = spread_of(design_stratified("region", n = p$n)),
        by_band   = spread_of(plan)))
```

`deff()` estimates that exchange rate from a single sample, which is exactly
what `plan_size(deff = )` wants when sizing next quarter's audit:

```{r replan}
plan_size(margin = target, sd = sd(stores$last_year), N = N,
          target = "total", response = 0.95,
          deff = deff(ht_total(audit, "loss")))
```

A design effect measured on one sample is itself an estimate, and the
simulation above puts the true ratio of variances somewhat higher. A cautious
planner rounds it up — but either way, the next audit can be a fraction of the
size of this one for the same precision.

## 6. Hand off

`drawn` estimates totals, means and proportions. For regression, calibration
or quantiles, hand the sample to `survey`; the design is expressed in `survey`'s
own terms, so the standard errors agree.

```{r survey, eval = requireNamespace("survey", quietly = TRUE)}
des <- as_svydesign(audit)
survey::svytotal(~loss, des)
```

## Where next

* `vignette("choosing-a-design")` compares every design on one population and
  shows when each pays.
* The articles on the package website explain how each variance estimator
  works, and walk through an audit sample and a spatially balanced field
  survey.
