---
title: "Walkthrough: an audit sample"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.show = "hold", dpi = 96,
                      fig.width = 7, fig.height = 3.4)
```

```{r setup, message = FALSE}
library(drawn)
```

An auditor has 2,000 invoices and time to examine about 150. The questions are
the usual ones: how much is the ledger misstated in total, with an upper bound
the auditor can sign off against; what share of invoices contain an error; and
is any site worse than the others? This walkthrough builds the sample in the
shape auditors use — examine every large item, sample the rest in proportion to
size — and shows what each choice buys.

## The ledger

Book values are known for every invoice. Misstatements are only found by
examining one. Here the ledger holds them too, so the answers can be checked,
but nothing below uses them except through the sample.

```{r ledger}
set.seed(31)
N <- 2000
ledger <- data.frame(
  invoice = sprintf("INV%05d", 1:N),
  site    = sample(c("north", "south", "east", "west"), N, replace = TRUE,
                   prob = c(0.35, 0.3, 0.2, 0.15)),
  book    = round(rlnorm(N, meanlog = 7, sdlog = 1.2), 2)
)
# Errors are more common at the west site; each misstates a share of the
# invoice's book value
error_rate <- c(north = 0.05, south = 0.05, east = 0.04, west = 0.15)
has_error <- runif(N) < error_rate[ledger$site]
ledger$misstatement <- ifelse(has_error,
                              round(ledger$book * runif(N, 0.05, 0.6), 2), 0)

c(book_total = sum(ledger$book), misstatement = sum(ledger$misstatement),
  error_rate = mean(ledger$misstatement > 0))
```

A handful of invoices carry a large share of the book value — the usual shape
of a ledger, and the reason simple random sampling is a poor fit:

```{r concentration}
top <- sort(ledger$book, decreasing = TRUE)
round(c(top_1pct = sum(top[1:20]), top_5pct = sum(top[1:100])) / sum(top), 2)
```

## Four ways to spend 150 examinations

* **Simple random**: every invoice equally likely.
* **Stratified by value band**, with Neyman allocation on book value: more
  examinations where values vary most.
* **Certainty plus PPS**: every invoice over 10,000 examined, and the rest drawn
  with probability proportional to book value — the design behind
  monetary-unit sampling.
* **Certainty plus spread**: the same certainty stratum, and the rest spread
  evenly along book value with equal probabilities — a fine stratification on
  size without choosing the band boundaries.

```{r designs}
ledger$band <- cut(ledger$book, c(0, 500, 2000, 10000, Inf),
                   labels = c("<500", "500-2k", "2k-10k", "10k+"))
n_certain <- sum(ledger$book >= 10000)
n_certain

designs <- list(
  simple    = design_simple(n = 150),
  banded    = design_stratified("band", n = 150, allocation = "neyman",
                                allocation_by = "book", min_per_stratum = 2),
  cert_pps  = design_certainty("book", 10000,
                               design_weighted("book", n = 150 - n_certain,
                                               method = "systematic")),
  cert_spread = design_certainty("book", 10000,
                                 design_spread("book", n = 150 - n_certain))
)
```

All four examine 150 invoices. Which is most precise for the misstatement total
depends on how misstatements relate to size — here they scale with book value —
and simulation settles it:

```{r compare}
spread_of <- function(design, R = 150) {
  sd(vapply(seq_len(R), function(i) {
    ht_total(draw(ledger, design, seed = i, weights = TRUE), "misstatement",
             variance = "none")$total
  }, numeric(1)))
}
sds <- vapply(designs, spread_of, numeric(1))
round(sds)
```

The designs that let book value decide how likely an invoice is to be examined
— Neyman bands, and certainty plus PPS — beat simple random sampling by
`r round(100 * (1 - min(sds[c("banded", "cert_pps")]) / sds[["simple"]]))`%
in standard error. That is a real gain but not a dramatic one, and the reason
is worth seeing: only `r round(100 * mean(ledger$misstatement > 0))`% of
invoices are misstated at all, so *which* invoices happen to contain an error
dominates the uncertainty whatever the design. Spreading the sample evenly along
book value with equal probabilities does not help here, because it does not
put more examinations where misstatements are largest.

The certainty-plus-PPS design is the one auditors know, and the one used
below.

## Draw and check

```{r draw}
audit_design <- designs$cert_pps
audit_design

s <- draw(ledger, audit_design, seed = 11, weights = TRUE)
sample_summary(s)
table(certain = s$.prob == 1)
```

## Estimate

Total misstatement, with a one-sided 95% upper bound — which is the upper end
of a two-sided 90% interval:

```{r total}
m <- ht_total(s, "misstatement", level = 0.90)
m
c(upper_bound_95 = m$ci[2], truth = sum(ledger$misstatement))
```

The variance is Deville's approximation for the PPS part; the certainty
invoices were all examined, so they add nothing to it.

The rate of invoices in error is the mean of a yes/no column. Its interval is
formed on the logit scale, so it stays inside 0 to 1 and is lopsided the way a
small proportion's uncertainty is:

```{r rate}
ht_mean(s, s$misstatement > 0)
```

The design effect is well above 1: a design built to estimate a total in
money is a poor one for counting errors, which do not scale with invoice size.
If the error rate mattered as much as the total, stratifying by value band
would serve both better.

And by site, from the same sample:

```{r by-site}
ht_mean(s, s$misstatement > 0, by = "site")
```

West stands out. Because site was not a stratum, the number of west invoices
examined was itself random, and the standard errors account for that.

## How far to trust the bound

Misstatements are rare and skewed: most examined invoices contribute zero, a
few contribute a lot. The normal and t approximations behind the interval
assume the estimate's sampling distribution is roughly symmetric, and that is
least true exactly here. So check it — draw the design many times and count how
often the upper bound actually sits above the true total:

```{r bound-coverage}
truth <- sum(ledger$misstatement)
covered <- vapply(1:300, function(i) {
  r <- ht_total(draw(ledger, audit_design, seed = 1000 + i, weights = TRUE),
                "misstatement", level = 0.90)
  r$ci[2] >= truth
}, logical(1))
mean(covered)
```

`r round(100 * mean(covered))`% — short of the 95% the bound claims. Most
samples contain only a few misstated invoices, so the estimate is usually a bit
low and occasionally far too high, and a symmetric interval around it is placed
too low exactly when it matters. That is not a defect of the variance
estimator; it is the normal approximation meeting a rare, skewed quantity.

Audit practice has bounds built for this case — the Stringer bound and
Bayesian methods, which the [`jfa`](https://cran.r-project.org/package=jfa)
package implements. Two things help within a design-based analysis: a larger
sample, which makes the estimate's distribution more symmetric, and a lower
one-sided level. `drawn`'s part is the design and a design-based estimate;
checking a bound this way before signing off against it is cheap, and this is
the case that shows why it is worth doing.

## Hand off

For anything further — a ratio of misstatement to book value by site, a
regression of taint on invoice characteristics — the sample goes to `survey`
with its design intact:

```{r survey, eval = requireNamespace("survey", quietly = TRUE)}
des <- as_svydesign(s)
survey::svyratio(~misstatement, ~book, des)
```
