---
title: "Walkthrough: a spatially balanced field survey"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.show = "hold", dpi = 96,
                      fig.width = 7.5, fig.height = 3.6)
```

```{r setup, message = FALSE}
library(drawn)
```

A forestry team needs total above-ground biomass for a 1,600-cell landscape
and can visit 60 plots. Biomass varies smoothly — it follows slope, soil and
stand history — so plots close together tend to agree. That single fact
decides the design: a sample that covers the landscape evenly learns far more
per plot than one that happens to bunch up.

## The landscape

Each cell has map coordinates, a forest type, and a satellite canopy-cover
index known everywhere before fieldwork. Biomass is only measured on the plots
visited; the frame holds it here so the answers can be checked.

```{r landscape}
set.seed(8)
grid <- expand.grid(x = 1:40, y = 1:40)
grid$x <- grid$x + runif(nrow(grid), -0.4, 0.4)
grid$y <- grid$y + runif(nrow(grid), -0.4, 0.4)
bump <- function(cx, cy, h, w) h * exp(-((grid$x - cx)^2 + (grid$y - cy)^2) / w)
grid$biomass <- 80 + 1.5 * grid$x + bump(10, 30, 120, 60) +
  bump(30, 12, 90, 90) + bump(28, 32, 60, 40) + rnorm(nrow(grid), 0, 12)
grid$canopy <- pmin(1, pmax(0.05, grid$biomass / 300 + rnorm(nrow(grid), 0, 0.08)))
grid$type <- ifelse(grid$x + grid$y > 45, "conifer", "broadleaf")

c(cells = nrow(grid), total_biomass = round(sum(grid$biomass)))
```

```{r map-truth, fig.width = 4.2, fig.height = 4, fig.alt = "A heat map of biomass over the 40 by 40 landscape, with three high patches and a gentle west-to-east rise."}
pal <- hcl.colors(30, "YlGn", rev = TRUE)
op <- par(mar = c(2, 2, 2, 1))
plot(grid$x, grid$y, pch = 15, cex = 0.9, asp = 1, axes = FALSE, ann = FALSE,
     col = pal[cut(grid$biomass, 30)])
title("biomass (the truth, unknown before fieldwork)", cex.main = 0.85,
      adj = 0, font.main = 1)
par(op)
```

## Simple random versus spread

`design_spread()` draws by the local pivotal method (Grafström, Lundström and
Schelin 2012). Every cell keeps the same inclusion probability, 60 / 1,600, but
a selected cell makes its neighbours less likely to be selected too, so the
sample covers the map.

```{r maps, fig.alt = "Two maps of the landscape. The simple random sample of 60 leaves gaps and clusters; the spread sample of 60 is evenly placed."}
srs <- design_simple(n = 60)
spread <- design_spread(c("x", "y"), n = 60, scale = FALSE)

op <- par(mfrow = c(1, 2))
plot(srs, grid, type = "map", coords = c("x", "y"), seed = 4,
     main = "simple random")
plot(spread, grid, type = "map", seed = 4, main = "spread")
par(op)
```

`scale = FALSE` because both coordinates are in the same unit; standardising
them would change what "near" means.

## What spreading is worth

Draw each design many times and look at the spread of the estimated totals:

```{r compare}
sd_of <- function(design, R = 100) {
  sd(vapply(seq_len(R), function(i) {
    ht_total(draw(grid, design, seed = i, weights = TRUE), "biomass",
             variance = "none")$total
  }, numeric(1)))
}
designs <- list(
  simple        = srs,
  spread        = spread,
  spread_canopy = design_spread(c("x", "y", "canopy"), n = 60),
  pps_canopy    = design_weighted("canopy", n = 60, method = "systematic"),
  spread_pps    = design_spread(c("x", "y"), n = 60, size = "canopy",
                                scale = FALSE)
)
sds <- vapply(designs, sd_of, numeric(1))
round(sds)
round((sds / sds[["simple"]])^2, 2)   # variance relative to simple random
```

Spreading over the map alone brings the variance down to
`r round(100 * (sds[["spread"]] / sds[["simple"]])^2)`% of simple random
sampling's, for the same 60 plots. The other rows show two ways of also using
the satellite index, which is known everywhere:

* **spread across coordinates and canopy** balances the sample on canopy cover
  as well as position — a stratification on both, without drawing a single
  boundary. With several columns `scale = TRUE` (the default) puts them on an
  equal footing.
* **probability proportional to canopy**, alone or combined with spreading,
  bets that canopy predicts biomass and visits dense stands more often. It
  beats simple random sampling, but here it does less than spreading alone:
  varying the probabilities adds weight variation, and biomass is not close
  enough to proportional to canopy to repay it. `deff()` from a pilot tells
  you which way that goes on your own landscape.

## One survey, start to finish

Take the spread design and draw the 60 plots the crew will visit:

```{r draw}
plots <- draw(grid, spread, seed = 3, weights = TRUE)
nrow(plots)
sample_summary(plots)
```

After fieldwork, the total, and a mean per hectare-cell:

```{r estimate}
ht_total(plots, "biomass")
c(truth = sum(grid$biomass))
```

The variance is the local mean estimator: each plot is compared with its three
nearest sampled neighbours, which is how a spatially balanced sample has to be
judged. A simple-random formula would give most of the gain back. (The seed
here was chosen to show a typical sample; the next section checks how often the
interval misses.)

Totals by forest type come from the same 60 plots:

```{r by-type}
ht_total(plots, "biomass", by = "type")
tapply(grid$biomass, grid$type, sum)
```

### Checking the interval

The local mean estimator is an approximation, documented as erring on the
conservative side. It is worth confirming on your own kind of landscape before
relying on it — here, 150 repeated surveys:

```{r coverage}
truth <- sum(grid$biomass)
out <- vapply(1:150, function(i) {
  r <- ht_total(draw(grid, spread, seed = 500 + i, weights = TRUE), "biomass")
  c(r$total, r$variance, r$ci[1] <= truth && truth <= r$ci[2])
}, numeric(3))
c(variance_ratio = mean(out[2, ]) / var(out[1, ]), coverage = mean(out[3, ]))
```

A variance ratio above 1 means the reported standard errors are, on average, a
little larger than the real ones; the interval covers at least as often as it
claims.

## Restricting to a region

Often only part of the landscape is in scope — one management unit, one
catchment. There are two ways to handle it, and they are not the same:

* `design_spatial()` takes an `sf` polygon and draws a *simple random* sample of
  the cells inside it. It decides where sampling happens, not how evenly.
* Filtering the frame to the region and then spreading gives an even sample of
  the region, and keeps the spreading gain.

```{r region, eval = requireNamespace("sf", quietly = TRUE)}
unit <- sf::st_sfc(sf::st_polygon(list(cbind(c(5, 25, 25, 5, 5),
                                             c(15, 15, 38, 38, 15)))))
unit <- sf::st_set_crs(unit, 3857)      # a planar CRS: coordinates in metres

in_unit <- design_spatial(c("x", "y"), region = unit, n = 30, crs = 3857)
sum(inclusion_prob(grid, in_unit) > 0)  # cells the design can reach

# The spread alternative: restrict the frame, then spread within it
unit_frame <- grid[inclusion_prob(grid, in_unit) > 0, ]
spread_unit <- design_spread(c("x", "y"), n = 30, scale = FALSE)

sd_total <- function(frame, design) sd(vapply(1:100, function(i) {
  ht_total(draw(frame, design, seed = i, weights = TRUE), "biomass",
           variance = "none")$total
}, numeric(1)))

round(c(simple_in_region = sd_total(grid, in_unit),
        spread_in_region = sd_total(unit_frame, spread_unit)))
```

Both estimate the total for the unit — `design_spatial()` gives every cell
outside it probability zero, so its Horvitz-Thompson total covers the inside
only — and the spread version does it with a smaller error.

## Where other tools fit

`drawn`'s spread design is the local pivotal method with a local-mean variance,
integrated with the same draw–estimate–hand-off workflow as every other design.
For more specialised spatial work:

* [`spsurvey`](https://cran.r-project.org/package=spsurvey) implements the
  generalized random tessellation stratified (GRTS) design (Stevens and Olsen
  2004) and a full analysis workflow for environmental monitoring, including
  designs over lines and areas.
* [`BalancedSampling`](https://cran.r-project.org/package=BalancedSampling)
  implements the pivotal and cube families in compiled code, including the
  local cube method, which balances on auxiliary totals exactly as well as
  spreading, and scales to much larger frames.

## References

Grafström, A., Lundström, N. L. P. and Schelin, L. (2012). Spatially balanced
sampling through the pivotal method. *Biometrics*, 68, 514–520.

Grafström, A. and Schelin, L. (2014). How to select representative samples.
*Scandinavian Journal of Statistics*, 41, 277–290.

Stevens, D. L. and Olsen, A. R. (2003). Variance estimation for spatially
balanced samples of environmental resources. *Environmetrics*, 14, 593–610.

Stevens, D. L. and Olsen, A. R. (2004). Spatially balanced sampling of natural
resources. *Journal of the American Statistical Association*, 99, 262–278.
